Profiles of Wind Direction and Speed in Turbulent Ekman Flow

### 2021-05-28 write/manuscript.tex
some edits on manuscript; bulk statistics for theory (mas trans[prt and turning); slight modification of theory to better match the super-geostrophic maximum in u

\documentclass[a4paper,11pt]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{natbib}

\newcommand{\p}{\partial}
\newcommand{\gzm}{{\xi}}%{\gamma z_\text{match}}}

\date{Draft \today}
\title{Towards a Universal Veering Profile for Turbulent Ekman Flow at arbitrary Reynolds number \\
  \normalsize LES and DNS of Turbulent Ekman Flow \\}

@@

\begin{abstract}
  An analytical formulation for the profiles of stream- and span-wise velocity in turbulent Ekman flow
  is developed based on asymptotic theory as well as data from direct numerical simulation and large-eddy simulation.
  %
  The profile of the stream-wise component follows the classical viscous, logarithmic and wake scaling.
  %
  The span-wise component poses a conceptual challenge to the channel-flow analogy
  in the context of asymptotic mixing, and it exhibits a mixed scaling in the surface layer, but follows
  outer scaling for most of the outer layer.
  %
  In both profiles, the wake region can be described by smooth error-function transition from a logarithmic layer
  to the constant outer velocity.

\end{abstract}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Introduction}

The exact veering of the wind away from its geostrophic direction is due to direct and indirect (due to turbulence) frictional effects,
and it plays a role for various practical problems: 
%
It sets the amount of cross-isobaric volume flow and is thus a key factor in the life-cycle of large-scale synoptic systems. This mechanism is commonly
referred to as 'Ekman pumping' referring to the perturbation of the geostrophic equilibrium that causes the veering as it was first described
and solved for analytically by \cite{Ekman:1905tl}.
%
On the mesoscale, shear in the surface layer is known to be crucial for the initiation and sustainment of strong convective events (REFERENCE?).
% 
And locally, directional sheer of the wind in the upper part of the surface layer may cause a yaw for large wind power generation devices (REFERENCE?).
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Formulation}
%
\paragraph{Physical Problem.} We consider incompressible, turbulent Ekman flow, i.e. the turbulent flow over a flat rotating plate, as
a physical model for the truly neutral atmospheric boundary layer (ABL).
%
That is, the f-plane approximation is used confining effects of rotation to the horizontal components of
velocity and neglecting dynamical effects due to latitudinal changes of the rate of rotation. 
%
%%PAR
%
\paragraph{Notation.} The dimensional velocity vector of the numerical simulations is $\underline{U} = (U_1,U_2,U_3) = (U,V,W)$ over the coordinate system
$Oxyz$, where an approximate alignment with the surface shear stress is achieved.
%
For analysis of the results, we use two coordinate systems that are
(i)  exactly aligned with the surface shear stress and labelled by an upper index $\alpha$ as in $\underline{U}^\alpha$ for the velocity vector, and
(ii) the coordinate system aligned with the free-atmosphere geostrophic wind labelled by an upper index $G$ as in $\underline{U}^G$.
%
We denote by $u_\star$ the modulus surface friction velocity and let $Z_\star=1/u_\star$ for brevity.

\section{Towards a Universal Velocity Profile for the Turbulent Ekman Layer}
\subsection{Drag-Law} 
The drag law for Ekman flow, including prediction of surface veering, has received considerable attention in the past.
\begin{itemize}
\item \cite{Rossby:1935wa} first approach;
\item \cite{Tennekes:1973jh, Blackadar:1968ew} asymptotic similarity;
\item \cite{Spalart:1989p2436, Spalart:2008p2432, Spalart:2009p2466} seminal LES studies in search for a universal log-law;
\item \cite{Ansorge:2014hf, Ansorge:2019di} re-evaluation of the constants with geophysical focus;
\item \cite{Hogstrom:1988wg, Hgstrm:1996kt} (re-)consideration of observational data.
  %

\end{itemize}
\begin{figure}
  \centerline{\includegraphics[width=0.7\textwidth]{../plot/ustar_alpha.pdf}}
  \caption{Variation of $Z_\star$ and surface veering with Reynolds number according to
    the theory by Spalart et al. (1989) and as estimated from DNS data}
  \label{fig:drag_law}
\end{figure} 
%
Attempts were also undertaken to obtain profiles of the wind speed: 
\cite{Gryning:2007dy} present an extension of the wind-speed profile beyond the surface layer using a neutral referece profile
and a stability correction;
\cite{Kelly:2010ev}, based on a probabilistic representation of stratification, develop a model for the long-term mean windspeed
in the atmospheric boundary layer and compare this with observation at different sites; 
\cite{Kelly:2016bs} demonstrate the effect of such improved model for wind-energy applications.
%
These investigations merely focus on the wind speed and thus do not regard the actual vector, i.e. the veering of
the wind with height is not described and there is little knowledge on the profile of the span-wise velocity component
and the precise shape of the hodograph in the limit of a truly neutral Ekman boundary layer.
%
While this is indeed an academic problem, the Ekman boundary layer often serves as limit for stratified boundary layers
and thus not only needs to be understood but also well-described as reference for higher-order approaches that take into
account possible effects of stratification, roughness or other physical complications of the geophysical system.
%
\subsection{Stream-wise Velocity} 
For the stream-wise velocity profile (that in non-rotating flows due to the geometry is always aligned with the surface shear stress),
well-established theories exist for various regimes according to their distance from the wall and the relative role of
viscosity, turbulence and interaction with the outer region of the flow with the logarithmic law for the mean velocity as a central anchor point:
\begin{subequations}
  \begin{itemize}
  \item In immediate vicinity to the surface, local turbulent mixing cannot occur for the no-slip/no-penetration boundary condition,
    and the mean velocity is described by a viscous profile of the form
    \begin{equation}
      (U,W)^{\alpha+}  = (z^+,0)  
    \end{equation}
    where the direction of 
    in absence of roughness elements are marginal linear regime / viscous sub-layer \cite{Foken:2002gv,Foken:1978bc} 
  \item buffer layer transition from linear regime to log-law  
  \item log-layer (K\'arm\'an, Prandtl, Jimenez, Marusic)
  \item outer log-layer () 
  \item wake region
  \item outer region 
  \end{itemize}
\end{subequations} 
%
\begin{figure}
  \includegraphics[width=\textwidth]{../plot/u_profile.pdf}
  \caption{Shear-aligned profiles of velocity components $U^{\alpha+}$ in inner (left) and outer (right) units.} 
\end{figure}

Existing theories applied to the boundary layer:
\begin{itemize}
\item \cite{Etling:2002kx, Emeis:2018vq}  (Sec. 21.10; Eq. 21.48) -- ad-hoc matching at Prandtl-layer height; requires $\alpha$ 
\item \cite{Emeis:2007cy} (Sec. 3; Eq. 3.1-3.19) -- 1D profile; considering constant (?) veering 
\item \cite{Gryning:2007dy} (Mean wind profiles)
\end{itemize}

For non-rotating, non-external flows theoretical considerations by Townsend (cite to attached-eddy hypothesis)
have recently been confirmed from both wind-tunnel  experiments and numerical simulation \cite{Ng:2011bk,Marusic:2013hf}
implying physical models for the second moments (Reynolds stresses) of the flow.

@@

The background rotation and associated veering of the surface wind implies a non-zero profile for the span-wise velocity that
challenges the conventional assumptions related to the so-called channel-flow analogy:

@@

This illustrates that the higher-order (in terms of the channel-flow analogy where $V=0$) correction follows a mixed scaling when the surface is approached which has been demonstrated for higher-order terms in other types of flow \citep{Mellado:2016hv}. 

@@

in the inner layer \citep{Marusic:2013hf} that also follow a mixed scaling in the inner layer.

@@

Based on the physical extent of the viscous sub-layer in Ekman flow around five wall units \citep{Foken:2002gv, Ansorge:2019di},

@@

\begin{figure}
  \phantom{AAA}\textbf{(a)\hspace{0.3\textwidth}(b)\hspace{0.3\textwidth}(c)}\\
  \includegraphics[width=\textwidth]{../plot/ekman_profiles_data.pdf}
  \caption{ \textbf{(a)} Velocity deficity, \textbf{(b)} velocity profile in shear-aligned hodographs and \textbf{(c)} hodograph in geostrophy-aligned coordinates.
    Thick, solid lines show theory, dashed lines data from DNS. }
  \label{fig:hodograph} 
\end{figure}

@@

\begin{subequations}
\begin{eqnarray}
  \left(\begin{matrix}
    \p_t U\\
    \p_t W  
  \end{matrix}\right)&=&\left(\begin{matrix}
     fW &+& \nu \p_z^2 U\\ 
    -f(U-G) &+& \nu \p_z^2 W
  \end{matrix}\right)\\ 
  \Rightarrow \partial_t (U+iW) &=& f(W-i(U-G))) + \nu \p_z^2(U+iW)
\end{eqnarray}
In stationary conditions, this system is solved by
\begin{eqnarray}
  \hat{u}(z) &=& U_{\inf} + e^{-\gamma z} \left[A \cos\gamma z + B \sin\gamma z\right]   \\
  \hat{w}(z) &=& W_{\inf} + e^{-\gamma z} \left[-A \sin\gamma z + B \cos\gamma z\right]
\end{eqnarray} 
\end{subequations}
where the constants $U_\inf$, $W_\inf$ set the top boundary condition and $A$ and $B$ set the bottom boundary condition. 
%
The most common boundary condition for a surface Ekman layer is $A=U_\inf=G$, $B=0$, and $W_\inf=0$.
%
The lower boundary condition, however, neglects the existence of the surface layer, and it appears reasonable to define
$A=c G$ where $c<1$ is a constant that icorporates the increased shear in the surface layer.
%
Given a 'matching height' $z_{match}$ and normalized matching height $\xi=\gamma z_\text{match}$ in the upper part of the inner layer, we can match the Ekman profile
to the inner layer by letting
%
\begin{eqnarray}
  \begin{matrix} 
  u(z_\text{match}) \equiv u_\text{match} &=& U_{\inf} + e^{-\gzm}} \left [ A \cos\gzm + B\sin\gzm \right] \\
    w(z_\text{match}) \equiv w_\text{match} &=& W_{\inf} + e^{-\gzm}} \left [-A \sin\gzm + B\cos\gzm \right]
  \end{matrix} 
\end{eqnarray}
\begin{eqnarray}
  e^\gzm  \left(\begin{matrix}
     u_\text{match}-U_{\inf} \\ 
     w_\text{match}-W_{\inf}
  \end{matrix}\right) = \left(\begin{matrix}
  \cos\gzm A &+ \sin\gzm B \\ -\sin\gzm A &+ \cos\gzm B
  \end{matrix} \right) \\
\end{eqnarray}
Matching the profile at $\xi=0$, one obtains
$A = \Delta u_\text{match} $ and $B=-\Delta w_\text{match} $. Otherwise, it is
\begin{subequations} 
\begin{eqnarray}
  e^\gzm \Delta u_\text{match} = A \cos\gzm + B\sin\gzm \\
  e^\gzm (\cos\gzm \Delta u_\text{match} - \sin\gzm \Delta w_\text{match}) = A (\cos^2\gzm + \sin^2\gzm)
\end{eqnarray}
\begin{eqnarray}
  \Rightarrow A &=&e^\gzm (\cos\gzm\Delta u_\text{match} + \sin\gzm \Delta w_\text{match})\\
  B&=&  \frac{e^\gzm \Delta u_\text{match} - A \cos \gzm}{\sin\gzm} \\
  &=& e^\gzm \frac{ \sin^2\gzm \Delta u_\text{match}+\sin\xi\cos\xi \Delta w_\text{match} }{\sin\xi }  \\
  &=& e^\gzm (\sin\xi\Delta u_\text{match} + \cos\xi\Delta w_\text{match}) 
\end{eqnarray}
\end{subequations} 
\bibliographystyle{plainnat}
\bibliography{ansorge.bib}

>>>>

\documentclass[a4paper,11pt]{amsart}

\usepackage[margin=3.5cm]{geometry}
\usepackage{lineno}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{xcolor}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% BIBLIOGRAPHY DETAILS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[style=authoryear-comp,firstinits=true,citestyle=authoryear,natbib=true,backend=bibtex]{biblatex}
\renewcommand\nameyeardelim{, }
\DeclareNameAlias{sortname}{family-given}
\renewcommand*{\multinamedelim}{\addcomma\space}
\renewcommand*{\finalnamedelim}{\addcomma\space}
\setlength{\bibitemsep}{\baselineskip}
\renewbibmacro{in:}{}

\AtEveryBibitem{%
  \clearfield{url}
  \clearfield{urlday}
  \clearfield{urlmonth}
  \clearfield{urlyear}
  \clearfield{issn}
  \clearfield{month}
%%%  \clearfield{doi}%
  \clearfield{keywords}
  \clearfield{note}
}

\bibliography{ansorge}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% CUSTOM DEFINITIONS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\p}{\partial}
\newcommand{\gzm}{{\xi}}%{\gamma z_\text{match}}}
\newcommand{\todo}[1]{\textcolor{red}{$[$#1$]$}}
\newcommand{\SFC}{\mathrm{sfc}}

\date{\scriptsize Draft - Version \today}
\title{Profiles of Wind Direction and Speed in Turbulent Ekman Flow}
\author{Cedrick Ansorge, Hauke Wurps}

@@

\begin{abstract}
  The profiles of wind speed and direction in turbulent Ekman flow
  are formulated based on asymptotic theory and data from direct numerical simulation. 
  %
  The profile of the stream-wise component follows the classical viscous, logarithmic and wake scaling.
  %
  In the outer layer, the velocity component profiles can be described by an Ekman-spiral with adapted
  boundary conditions that result in a reduction of the spiral-like rotation. 
  %
  The span-wise component poses a conceptual challenge to the channel-flow analogy
  in the context of asymptotic matching; it exhibits a mixed scaling in the surface layer, but follows
  outer scaling for most of the outer layer.
  %
\end{abstract}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Introduction}

Wind veer away from its geostrophic direction is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
physical effects on large and small scales.
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the mesoscale, shear in the surface layer a crucial factor for the initiation and sustainment of strong convective events (REFERENCE?).
% 
Locally, the directional shear of the wind in the upper part of the surface layer may cause a systmatic yaw
for tall wind power generation devices where blades reach into the Ekman layer, i.e. that part of the boundary
layer where the wind starts to turn. 
%
\par
%
Wind veer in the planetary boundary layer is commonly characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where $u_\star\equiv\sqrt{\nu|\tau_\SFC|}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the alignment of the frame of reference.
%
Therefore, knowing these parameters is a pre-requisite for any quantitative theory of the surface layer and a first
attempt in theoretical quantification was undertaken by \cite{rossby:PPO1935} using integral relations in the boundary layer.
%
Asymptotic similarity theory was later used by \cite{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the atmospheric boundary layer based on
observations \cite{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}. 
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height;
\cite{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48) choose the Prandtl-layer height to match
the wind speed profiles, which, however, requires to externally prescribe the veering $\alpha$ at that height. 
A one-dimensional profile with constant veering is given by \cite[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}.

\cite{gryning:BM2007} present an extension of the wind-speed profile beyond the surface layer
using a neutral referece profile and a stability correction;
\cite{kelly:BM2010}, based on a probabilistic representation of stratification,
develop a model for the long-term mean windspeed
in the atmospheric boundary layer and compare this with observation at different sites; 
\cite{kelly:WE2016} demonstrate the effect of such improved model for wind-energy applications.
In consideration of the large scale separation in geophysical flow, the rotation of the wind in the
surface layer is often assumed negligible, and above investigations merely focus on the wind speed; 
that means, the veering of the wind with height is not described and there is little knowledge on the
profile of the span-wise velocity component and the precise shape of the hodograph in the limit of a
truly neutral Ekman boundary layer.
%%% \par  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% % No discussion of higher-order moments at this stage
%%%
%%% For non-rotating, non-external flows theoretical considerations by Townsend (cite to attached-eddy hypothesis)
%%% have recently been confirmed from both wind-tunnel  experiments and numerical simulation
%%% \cite{ng_comparison_2011,marusic_logarithmic_2013};
%%% this implies that physical models for the second moments (Reynolds stresses) of the flow \ldots.
\par %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Ekman-layer models and two-layer models of the ABL take into account rotational effects at higher altitudes,
for instance when the wind speed needs to be evaluated at heights on the order of 100-200\,\mathrm{m},
a particular concern when it comes to wind-power forecasting \citep{optis:BM2014}.
Despite rotational effects being considered, the formulation of these models for the outer layer
and analysis of their performance primarily focuses on wind speed.
\cite{jiang:JAS2018} recognized that the outer part of the Ekman boundary layer receives less attention in comparison
with the surface layer and commence in studying the neutral problem by Large-Eddy simulation (LES).
They focus on the wind speed and find an extended logarithmic layer when considering the wind speed instead of
the shear-aligned component, and they eventually demonstrate by means of an analytical model that this vertical
extension of the logarithmic layer may be explained by a transfer of stress to the spanwise velocity component where
it is assumed that the shear vector $\tau(z)$ and stress vectors $(\p_zU,\p_zV)$ are aligned. 
Turbulent Ekman flow is considered here as a conceptual model of the
homogeneous, stationary atmospheric boundary layer over a flat surface under neutral stratification. 
Universal profiles of the wind vector for turbulent Ekman flow can serve as well-described limit and reference
for higher-order approaches taking into account possible effects of stratification, roughness or other physical
complications encountered in the realgeophysical system.
While, on first sight, the study of such a strongly idealized case appears as an academic problem, it contains the essence
of surface similarity as it is used in most atmospheric models, be it conceptual or numeric ones.
For instance, effects of roughness are commonly incorporated by a linear transformation of vertical scale involving the
roughness parameter $z_0$ and for larger roughness also a displacement height~%
\citep{monin:1975,jacobs:AFM1988,hogstrom:BM1988}; 
effects of stability are commonly accounted for by a linearization around the neutrally stratified profile~%
\citep{monin:ARF1970, monin:1975, hogstrom:BM1988, hogstrom:BM1996,sakagami:BM2020}.
Such generalized profiles may also serve as better initial condition for numerical simulation of the flow,
to minimize the length of initial transient periods, or as benchmark for turbulence
closures that can be tuned to reproduce the neutral limit case.
Despite the strong simplifications implied by our choice of set-up, there is no straightforward approach to solving
this conceptually well-defined problem.
Large-Eddy simulation not only needs to be tuned for the surface shear stress and veering angle, but it also
relies on sub-grid closures that often assume alignment of the stress with gradients -- a questionable pre-requisite when the wind rotates with height.
While recent advances have been made in analysis of this simplified set-up \cite{jiang:JAS2018} (MORE?),
there is not enough understanding for a quantitative generalization of the results to arbitrary external forcing.
At the same time, an increasing amount of high-quality and high-resolution data from turbulence-resolving approaches
is emerging due to recent advances in high-performance computing and its application to geophysical problem sets;
the geophysical range of scale separation, however, is--and it will remain so for the foreseeable future--out of reach
for such simulation.
Here, the routinely employed concept of Reynolds-number similarity can help: it states that certain statistics (normally those of low order) become independent of the Reynolds number when scaled appropriately \todo{Dimotakis / da Silva / Pope referece?} .
It appears that for certain statistics in Ekman flow, this regime is reached with the Reynolds numbers that became
possible with the increasing capabilities of supercomputering over the past years.
The robust features of mean velocity profiles across a range of Reynolds numbers
are exploited in this work to develop a formulation for
both the stream-wise and span-wise components of the mean velocity vector as a function of the
Reynolds number.

\todo{Numerical approach, DNS, Scale-separation, Re-similarity
  \citep{dimotakis:JFM2000, dasilva:ARF2014, pope:2000, moin:ARF1998, mellado:JAM2018}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Formulation}
%
We consider here incompressible, turbulent Ekman flow, that is, the turbulent flow over a flat rotating plate, as
a physical model for the truly neutral atmospheric boundary layer (ABL).
%
The f-plane approximation is applied such that rotation only acts on horizontal velocity components;
we thus neglect rotational effects on the horizontal components of velocity and dynamical effects
due to latitudinal variation of the rate of rotation.
%
\par
%
The dimensional velocity vector of the numerical simulations is $\underline{U} = (U_1,U_2,U_3) = (U,V,W)$ over the coordinate system
$Oxyz$, where an approximate alignment (plus/ minus few degrees centigrade) of the direction $Ox$ with the surface shear stress is achieved.
The coordinate $Oy$ points away from the wall,  and $Oz$ points in the span-wise direction normal to $Oxy$. 
%
For analysis of the results, we use two coordinate systems that are
(i)  exactly aligned with the surface shear stress and labelled by an upper index $\alpha$ as in $\underline{U}^\alpha$ for the velocity vector, and
(ii) the coordinate system aligned with the free-atmosphere geostrophic wind labelled by an upper index $G$ as in $\underline{U}^G$.
%
We denote by $u_\star$ the modulus surface friction velocity and let $Z_\star=1/u_\star$ for brevity.

\section{Towards a Universal Velocity Profile for the Turbulent Ekman Layer}

Formulations for the outer layer that take into account the rotation, and thus deviation from the channel-flow analogy
need to be matched to the framework of surface similarity, and it
occurs that a smooth transition from the inner to the outer layer is not easily achieved.
%
\cite{optis:BM2014}, for instance, define an ``effective geostrophic wind vector that has the same magnitude of the
observed surface geostrophic wind and is rotated by the angle $\alpha$ [their nomenclature]'' to overcome the
unsteady transition when coming approaching the Ekman layer from below.
%
While such rotation of the wind vector is \emph{a posteriori} justified by the observational data that the model outcomes are compared to,
we believe that it is the manifestation of a mismatch between the theoretical treatment of the inner and outer layer of
turbulent Ekman flow.

\end{itemize}
\begin{figure}
  \centerline{\includegraphics[width=0.7\textwidth]{../plot/ustar_alpha.pdf}}
  \caption{Variation of $Z_\star$ and surface veering with Reynolds number according to
    the theory by Spalart et al. (1989) and as estimated from DNS data}
  \label{fig:drag_law}
\end{figure}
% 
\subsection{Drag-Law}
A drag-law for Ekman flow determines two key parameters of the flow as a function of Reynolds number alone.
%
These are, first, the modulus of the surface friction normalized by the geostrophic wind forcing, $u_\star$ and,
second, the direction of surface shear stress, $\alpha$, also termed wind veering. 
%
A non-zero veering of the wind is a rather special case in comparison with most turbulent flows conisdered in an
engineering context, and it confronts us with a situation where the most appropriate coordinate system for analysis
(namely that aligned with the surface shear stress) is a priori unknown. 
%
We compare our DNS data against a semi-empirical drag-law based on integral consideration \cite{spalart:JFM1989}
and find, as demonstrated in previous work \cite{ansorge:BM2014}, excellent agreement in the range
 $400<\mathbf{Re}<1600$, representing a factor of 16 in variation of viscosity. 
%
\par
%
We also find that the solution of the transient equation involved in estimation of $u_\star$ for a given Reynolds number
$Re_D$ is approximated reasonably by the formulation
\begin{equation}
  u_\star= 4\log(\mathbf{Re}_D))-8
\end{equation}
which quantifies the 'weak' dependence of $u_\star$ on the Reynolds number as an approximately logarithmic one, at least
for problems with a scale separation on the order that is relevant to geophysical problem ($Re_D<10^{8}$).  
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection{Profile in the outer layer}
%
The outer layer of Ekman flow is characterized by a turning of the wind velocity and the super-geostrophic maximum
that is sustained by momentum convergence at the inflection point of the velocity profile. 
The super-geostrophic maximum of streamwise velocity and a secondary minimum aloft the bulk-turbulent part of
the boundary layer are well-described by a classic Ekman spiral with adapted boundary conditions and a shift
in reference height.
%
The textbook solution for the Ekman layer reads as 
\todo{equations / figures}
\begin{figure}
  \includegraphics[width=\textwidth]{../plot/outer_layer.pdf}
  \caption{} 
\end{figure}

\subsection{Stream-wise Velocity Component} 
For the stream-wise velocity profile (that in non-rotating flows due to the geometry is always aligned with the surface shear stress),
well-established theories exist for various regimes according to their distance from the wall and the relative role of
viscosity, turbulence and interaction with the outer region of the flow with the logarithmic law for the mean velocity as a central anchor point.
%
\par
%
\begin{subequations}

In immediate vicinity to the surface, local turbulent mixing cannot occur for the no-slip/no-penetration boundary condition,
and the mean velocity is described by a viscous profile of the form
\begin{equation}
  (U,W)^{\alpha+}  = (z^+,0)  
\end{equation}
where the direction of the velocity points into the exact opposite direction of the wall shear stress $\tau$. 
In absence of roughness elements and for small roughness ($z_0^+<5$), this linear regime is known as
viscous sub-layer \cite{foken:m2002,foken:BM1978}.
In fact, this law of the wall has no degree of freedom given the drag, i.e. once $u_\star$ and $\alpha$ are defined. 
However, theoretical foundation is lacking for the exact shape of the velocity profile in the buffer layer; though crucial for turbulence
production, it is commonly understood as a transition region between the linear profile at the surface and the logarithmic profile
aloft.
A pure blending from the linear velocity profile into the logarithmic one is, however, not reasonable as both the linear and logarithmic
profile overestimate the velocity in the buffer layer. We therefore introduce a higher-order compensating term into
the linear profile before blending it into the logarithmic law:
\begin{equation}
  U_\text{inner}^{\alpha}^+ = \frac{ y^+ + \gamma_4 (y^+)^{4} + \gamma_6 (y^+)^{6} } { 1 + \gamma_6 /u_{ref} (y^+ )^6}, 
\end{equation} 
where the fourth- and sixth-order terms in the nominator are designed to fit the profile below $y^+\approx 10$ and
the sixth-order correction in the denominator limits the growth of $u_\text{inner}$ for large $y^+$.
In the logarithmic region, we use the profile
\begin{equation}
  U_\text{log}^{\alpha}^+ = \frac{1}{\kappa} \log y^+ + C 
\end{equation}
with the von-K\'arm\'an constant $\kappa=0.416$ and the boundary condition $C=5.4605$. %  log-layer (K\'arm\'an, Prandtl, Jimenez, Marusic)
\par %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \includegraphics[width=\textwidth]{../plot/u_profile.pdf}
  \caption{Shear-aligned profiles of velocity components $U^{\alpha+}$ in inner (left) and outer (right) units.} 
\end{figure} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The background rotation and associated veering of the surface wind implies a non-zero profile for the span-wise
velocity which challenges the conventional assumptions related to the so-called channel-flow analogy:

@@

This illustrates that the higher-order (in terms of the channel-flow analogy where $V=0$) correction follows a mixed scaling when the surface is
approached which has been demonstrated for higher-order terms in other types of flow \citep{mellado:BM2016}. 

@@

in the inner layer \citep{marusic:JFM2013} that also follow a mixed scaling in the inner layer.

@@

Based on the physical extent of the viscous sub-layer in Ekman flow around five wall units \citep{foken:m2002, ansorge:BM2019},

@@

\begin{figure}
  \phantom{AAA}\textbf{(a)\hspace{0.3\textwidth}(b)\hspace{0.3\textwidth}(c)}\\
  \includegraphics[width=\textwidth]{../plot/ekman_profiles.pdf}
  \caption{ \textbf{(a)} Velocity deficity, \textbf{(b)} velocity profile in shear-aligned hodographs and \textbf{(c)} hodograph in geostrophy-aligned coordinates.
    Thick, solid lines show theory, dashed lines data from DNS. }
  \label{fig:hodograph} 
\end{figure}

@@

\begin{subequations}
\begin{eqnarray}
  \left(\begin{matrix}
    \p_t U\\
    \p_t W  
  \end{matrix}\right)&=&\left(\begin{matrix}
     fW &+& \nu \p_z^2 U\\ 
    -f(U-G) &+& \nu \p_z^2 W
  \end{matrix}\right)\\ 
  \Rightarrow \partial_t (U+iW) &=& f(W-i(U-G))) + \nu \p_z^2(U+iW)
\end{eqnarray}
In stationary conditions, this system is solved by
\begin{eqnarray}
  \hat{u}(z) &=& U_{\infty} + e^{-\gamma z} \left[A \cos\gamma z + B \sin\gamma z\right]   \\
  \hat{w}(z) &=& W_{\infty} + e^{-\gamma z} \left[-A \sin\gamma z + B \cos\gamma z\right]
\end{eqnarray} 
\end{subequations}
where the constants $U_\infty$, $W_\infty$ set the top boundary condition and $A$ and $B$ set the bottom boundary condition. 
%
The most common boundary condition for a surface Ekman layer is $A=U_{\infty}=G$, $B=0$, and $W_{\infty}=0$.
%
The lower boundary condition, however, neglects the existence of the surface layer, and it appears reasonable to define
$A=c G$ where $c<1$ is a constant that icorporates the increased shear in the surface layer.
%
Given a 'matching height' $z_{match}$ and normalized matching height $\xi=\gamma z_\text{match}$ in the upper part of the inner layer, we can match the Ekman profile
to the inner layer by letting
%
\begin{eqnarray}
  \begin{matrix} 
    u(z_\text{match}) \equiv u_\text{match} &=& U_{\infty} + e^{-\gzm}} \left [ A \cos\gzm + B\sin\gzm \right] \\
    w(z_\text{match}) \equiv w_\text{match} &=& W_{\infty} + e^{-\gzm}} \left [-A \sin\gzm + B\cos\gzm \right]
  \end{matrix} 
\end{eqnarray}
\begin{eqnarray}
  \Rightarrow \left(\begin{matrix}
    u_\text{match}-U_{\infty} \\ 
    w_\text{match}-W_{\infty}
  \end{matrix}\right) = e^{-\gzm}\left(\begin{matrix}A \\ B \end{matrix}\right)\left(\begin{matrix}
    \cos\gzm &+ \sin\gzm \\ -\sin\gzm &+ \cos\gzm
  \end{matrix} \right) \\
\end{eqnarray}
Matching the profile at $\xi=0$, one obtains
$A = \Delta u_\text{match} $ and $B=-\Delta w_\text{match} $. Otherwise, it is
\begin{subequations} 
\begin{eqnarray}
  e^\gzm \Delta u_\text{match} = A \cos\gzm + B\sin\gzm \\
  e^\gzm (\cos\gzm \Delta u_\text{match} - \sin\gzm \Delta w_\text{match}) = A (\cos^2\gzm + \sin^2\gzm)
\end{eqnarray}
\begin{eqnarray}
  \Rightarrow A &=&e^\gzm (\cos\gzm\Delta u_\text{match} + \sin\gzm \Delta w_\text{match})\\
  B&=&  \frac{e^\gzm \Delta u_\text{match} - A \cos \gzm}{\sin\gzm} \\
  &=& e^\gzm \frac{ \sin^2\gzm \Delta u_\text{match}+\sin\xi\cos\xi \Delta w_\text{match} }{\sin\xi }  \\
  &=& e^\gzm (\sin\xi\Delta u_\text{match} + \cos\xi\Delta w_\text{match}) 
\end{eqnarray}
\end{subequations}

\printbibliography
